---
title: "Fibrinogen - Haptoglobin: DESeq2 analysis of KOS"
author: "Gabriela Puerres"
date: "`r Sys.Date()`"
output: 
  html_document:
    toc: true
    toc_float:
      collapsed: true
      smooth_scroll: true
    theme: cosmo
---

```{r setup, include=FALSE}
# R Markdown setup chunk
# Define your base paths
base_dir   <- path.expand("C:/Users/diana/OneDrive - purdue.edu/Master/Slivozkyi_Lab/Projects/Cow_Inflammation_Study/PICRUst2/DESeq2/Fibrinogen_Haptoglobin_Final/")
cache_dir  <- paste0(file.path(base_dir, "Files/Cache"), "/")
fig_dir    <- paste0(file.path(base_dir, "Files/Figures/"), "/")

# Create directories if they don't already exist
dir.create(cache_dir, recursive = TRUE, showWarnings = FALSE)
dir.create(fig_dir,   recursive = TRUE, showWarnings = FALSE)

# Set knitr chunk options
knitr::opts_chunk$set(
  fig.keep      = "all",
  fig.path      = fig_dir,  
  cache         = TRUE,
  cache.lazy    = FALSE,
  cache.comments= FALSE,
  cache.path    = cache_dir)  # ends with a slash
```

# Input Data & Metadata

```{r data, message=FALSE, warning=FALSE}
# packages
library(phyloseq)
library(DESeq2)
library(tidyverse)
library(pheatmap)
library(ggplot2)
library(apeglm)
library(reactable)
library(DT)

# Working directory:
setwd("C:/Users/diana/OneDrive - purdue.edu/Master/Slivozkyi_Lab/Projects/Cow_Inflammation_Study/PICRUst2/")

# files: PICRUSt2 table ================================
picrust_file <- "Results/KO_metagenome_out/pred_metagenome_unstrat.tsv/pred_metagenome_unstrat.tsv"
data <- read.delim(picrust_file, check.names = FALSE, stringsAsFactors = FALSE)
remove_samples <- c("mock-synthetic-ZYMO-1",
                    "mock-synthetic-ZYMO-2",
                    "mock-synthetic-ZYMO-3",
                    "mock-synthetic-ZYMO-4",
                    "mock-synthetic-ZYMO-5")

data <- data[, !colnames(data) %in% remove_samples]
dim(data)
datatable(data, options = list(
dom = 'Bfrtip', # show buttons, filter, table
pageLength = 10,
lengthMenu = c(10, 15, 25, 50, 100),
scrollX = TRUE,
scrollY = "400px"))

# assume first column is feature IDs: KO
feat_col <- colnames(data)[1]
rownames(data) <- data[[feat_col]]
data[[feat_col]] <- NULL
data <- as.matrix(data) 


# files: meta table ================================

meta_file <- "C:/Users/diana/OneDrive - purdue.edu/Master/Slivozkyi_Lab/Projects/Cow_Inflammation_Study/PICRUst2/Updated_Cow_Metadata.csv"
metadata_clean <- read.csv(meta_file, header = TRUE, row.names = 1)
metadata_clean <- metadata_clean %>%  filter(Sample_Type == "Feces")
all(colnames(data) %in% rownames(metadata_clean))
# Subset metadata
covariates <- c("Sample_ID", "ExtractionDate","DIM_Collected", 
 "Parity", "BCS", "Fibrinogen_InflamStat",  "Haptoglobin_InflamStat", "qpcr_16s_copies_ul", "nonchim")

metadata <- metadata_clean[, covariates]

# Re-define parity: Uniparous (Parity = 1) vs Multiparous (Parity > 1).
metadata$Parity_Classification <- ifelse(metadata$Parity == 1,
                                         "Uniparous",
                                         "Multiparous")


# Convert all to factors
metadata$ExtractionDate <-factor(metadata$ExtractionDate)
metadata$DIM_Collected <- factor(metadata$DIM_Collected)
metadata$Parity_Classification <- factor(metadata$Parity_Classification)
metadata$BCS <- factor(metadata$BCS)
metadata$Fibrinogen_InflamStat <- factor(metadata$Fibrinogen_InflamStat)
metadata$Haptoglobin_InflamStat <- factor(metadata$Haptoglobin_InflamStat)

# for copy number
metadata$qpcr_16s_copies_ul <-  gsub(",", "", metadata$qpcr_16s_copies_ul)
metadata$qpcr_16s_copies_ul <- as.numeric(metadata$qpcr_16s_copies_ul)
metadata$qpcr_16s_copies_ul <- scale(metadata$qpcr_16s_copies_ul, center = T, scale = T) #mean-centered, variance-standardized f

#for nochim
metadata$nonchim <- as.numeric(metadata$nonchim)
metadata$nonchim <- scale(metadata$nonchim, center =T, scale = T)
```

# Fibrinogen Model

The model takes into account the effect of Parity, Days in Milk (DIM), Body Condition Score (BCS) + 16s qPCR Copy Number (qpcr_16s_copies_ul).

1.  Data set for DESeq2:

```{r DESeqDataSetFromMatrix}
library("DESeq2")

# Convert PICRUSt2 output to integer table
countmat <- round(data)

# DESeq2 object
dds <- DESeqDataSetFromMatrix(countData = countmat,
                              colData = metadata,
                              design = ~ Fibrinogen_InflamStat)
dds

# estimates size factors that normalize the count data to account for differences in sequencing depth (library size) or total counts per sample: 
dds <- estimateSizeFactors(dds, type = "poscounts")
head(sizeFactors(dds))
```

2.  Filtration of KOs. Prevalence cutoff of 5%

```{r}
# keeps only those KOs that are expressed above the prevalence cutoff:
# Set prevalence cutoff threshold:
prevalence_cutoff <- 0.05
counts_mat <- counts(dds)
prevalence <- rowSums(counts_mat > 1) / ncol(counts_mat)
keep <- prevalence >= prevalence_cutoff
dds_filtered <- dds[keep, ]
dds_filtered
```

3.  Run DESeq2:

```{r}
#  copy of the DESeqDataSet
dds_fb <- dds_filtered 

# using relevel, just specifying the reference level:
dds_fb$Fibrinogen_InflamStat <- relevel(dds_fb$Fibrinogen_InflamStat, ref = "Normal")

# re-run DESeq:
design(dds_fb) <- formula(~ Fibrinogen_InflamStat + Parity_Classification + DIM_Collected + BCS + qpcr_16s_copies_ul)

dds_fb <- DESeq(dds_fb, 
                 test = "Wald", 
                 fitType = "parametric")
```

4.  DESeq Results and Log2 Fold Change Shrinkage

```{r}
alpha <- 0.01
resultsNames(dds_fb)

res_fb <- results(dds_fb,
                   alpha = alpha,
                   name = "Fibrinogen_InflamStat_Elevated_vs_Normal")
summary(res_fb)

# Log2 Fold Change Shrink
res_LFC_fb <- lfcShrink(dds_fb, 
                         type = "ashr", 
                         res = res_fb)

# Filter for significant results
res_LFC_fb <- res_LFC_fb[which(res_LFC_fb$padj < 0.01 & 
                               abs(res_LFC_fb$log2FoldChange) > 2), ]
summary(res_LFC_fb)

# Order LFC results by adjusted P values
res_LFC_fb_df <- as.data.frame(res_LFC_fb[order(res_LFC_fb$padj, na.last = NA), ])
sum(res_LFC_fb_df$padj < 0.01, na.rm = TRUE)

# Keep only KOs with adjusted p-value < 0.01
res_LFC_fb_df <- res_LFC_fb_df[
  res_LFC_fb_df$padj < 0.01 & !is.na(res_LFC_fb_df$padj), ]
dim(res_LFC_fb_df)

# Save File
write.csv(res_LFC_fb, "DESeq2/Files/Results/Fibrinogen_full_model.csv")

# Plots
plotMA(res_fb, ylim = c(-30, 30))  # after DESeq2
plotMA(res_LFC_fb, ylim = c(-30, 30))  # after Log2 Fold Change shrinkage
```

5.  Volcano plot of significant p adjusted values

```{r Volcano_plot}
library(ggplot2)
library(ggrepel)

# Add threshold labels
res_df <- res_LFC_fb_df
res_df$significant <- ifelse(res_df$padj < 0.01 & abs(res_df$log2FoldChange) > 1,
                             "Significant", "Not significant")

# 10 top genes
top_genes <- res_df[which(res_df$padj < 0.01 & abs(res_df$log2FoldChange) > 2), ]

# Volcano plot
ggplot(res_df, aes(x = log2FoldChange, y = -log10(padj))) +
  geom_point(aes(color = significant), alpha = 0.7, size = 2) +
  scale_color_manual(values = c("Significant" = "red", "Not significant" = "gray")) +
  geom_vline(xintercept = c(-1, 1), linetype = "dashed", color = "black") +
  geom_hline(yintercept = -log10(0.01), linetype = "dashed", color = "black") +
  labs(title = "Fibrinogen KOs -Volcano Plot",
       x = expression(Log[2]~Fold~Change),
       y = expression(-Log[10]~Adjusted~p~value)) + geom_text_repel(data = top_genes,
                  aes(label = rownames(top_genes)),
                  size = 3, box.padding = 0.3, max.overlaps = 10) + theme_minimal(base_size = 14)


```

```{r Fb_Significant_KOs_Table}
# Display table
head(res_LFC_fb_df)

library(reactable)
library(DT)

res_df <- res_LFC_fb_df
if (!"KOs" %in% colnames(res_df)) {
res_df <- cbind(KOs = rownames(res_df), res_df)
rownames(res_df) <- NULL
}

datatable(
res_df,
extensions = c('Buttons', 'Responsive', 'Scroller'),
options = list(
dom = 'Bfrtip', # show buttons, filter, table
buttons = c('copy', 'csv', 'excel', 'pdf'),
pageLength = 15,
lengthMenu = c(10, 15, 25, 50, 100),
scrollX = TRUE,
scrollY = "400px",
deferRender = TRUE,
scroller = TRUE
),
filter = 'top',
rownames = FALSE,
class = 'stripe hover compact'
)
```

```{r fb_Anotation}
library(KEGGREST)

# KOs names:
KOs_names_Fibrinogen <- rownames(res_LFC_fb_df)
KOs_names_Fibrinogen <- gsub("ko:", "", KOs_names_Fibrinogen)
res_LFC_fb_df$X <- KOs_names_Fibrinogen


# Retrieve pathways for each KO
names(KOs_names_Fibrinogen) <- KOs_names_Fibrinogen

Fibrinogen_info <- lapply(KOs_names_Fibrinogen, function(ko) {
  ko_entry <- tryCatch(keggGet(paste0("ko:", ko))[[1]], error = function(e) NULL)
  ec_name <- if (!is.null(ko_entry) && !is.null(ko_entry$NAME)) ko_entry$NAME[1] else NA
  
  res <- tryCatch(keggLink("pathway", paste0("ko:", ko)), error = function(e) NULL)
  if (!is.null(res) && length(res) > 0) {
    path_ids <- sub("path:", "", res)
    path_names <- sapply(path_ids, function(pid)
      tryCatch(keggGet(pid)[[1]]$NAME, error = function(e) NA))
    list(EC_name = ec_name, Pathway = unique(na.omit(path_names)))
  } else {
    list(EC_name = ec_name, Pathway = NA)
  }
})

# Convert list to dataframe
Fibrinogen_info_df <- do.call(rbind, lapply(Fibrinogen_info, function(x) {
  # If multiple pathways, collapse into a single string separated by "; "
  pathway <- if (is.na(x$Pathway[1])) NA else paste(x$Pathway, collapse = "; ")
  data.frame(EC_name = x$EC_name, Pathway = pathway, stringsAsFactors = FALSE)
}))

#Update Data frame:
rownames(Fibrinogen_info_df) <- names(KOs_names_Fibrinogen)

Fibrinogen_merged <- cbind(res_LFC_fb_df,
                            Fibrinogen_info_df[res_LFC_fb_df$X, ])

write.csv(Fibrinogen_merged, "DESeq2/Files/Results/Fibrinogen_with_KEGG_info.csv")
```

```{r annotation_fb}
head(Fibrinogen_merged)
dim(Fibrinogen_merged)

head(rownames(Fibrinogen_merged))
# Annotated table

res_df_fb <- Fibrinogen_merged
res_df_fb$Pathway <- enc2utf8(as.character(res_df_fb$Pathway))
res_df_fb$EC_name <- enc2utf8(as.character(res_df_fb$EC_name))
```

```{r datatable_fb}
colnames(res_df_fb)
head(rownames(res_df_fb))
if (!"KOs" %in% colnames(res_df_fb)) {
  res_df_fb <- cbind(KOs = rownames(res_df_fb), res_df_fb)
  rownames(res_df_fb) <- NULL}


head(res_df_fb)

library(DT)
# Dataframe
DT::datatable(
  res_df_fb,
  extensions = 'Buttons',
  options = list(
    dom = 'Bfrtip',
    buttons = c('copy', 'csv', 'excel', 'pdf'),
    pageLength = 10,
    lengthMenu = c(10, 15, 25, 50, 100),
    scrollX = TRUE
  ),
  filter = 'top',
  rownames = FALSE,
  class = 'stripe hover compact')
```

# Haptoglobin Model

The model takes into account the effect of Parity, Days in Milk (DIM), Body Condition Score (BCS) + 16s qPCR Copy Number (qpcr_16s_copies_ul).

1.  Data set for DESeq2:

```{r Hp_DESeqDataSetFromMatrix}
library("DESeq2")

# Convert PICRUSt2 output to integer table
countmat <- round(data)

# DESeq2 object
dds <- DESeqDataSetFromMatrix(countData = countmat,
                              colData = metadata,
                              design = ~ Haptoglobin_InflamStat)
dds

# estimates size factors that normalize the count data to account for differences in sequencing depth (library size) or total counts per sample: 
dds <- estimateSizeFactors(dds, type = "poscounts")
head(sizeFactors(dds))
```

2.  Filtration of KOs. Prevalence cutoff of 5%

```{r filtration}
# keeps only those KOs that are expressed above the prevalence cutoff:
# Set prevalence cutoff threshold:
prevalence_cutoff <- 0.05
counts_mat <- counts(dds)
prevalence <- rowSums(counts_mat > 1) / ncol(counts_mat)
keep <- prevalence >= prevalence_cutoff
dds_filtered <- dds[keep, ]
dds_filtered
```

3.  Run DESeq2:

```{r DESeq}
#  copy of the DESeqDataSet
dds_hp <- dds_filtered 

# using relevel, just specifying the reference level:
dds_hp$Haptoglobin_InflamStat <- relevel(dds_hp$Haptoglobin_InflamStat, ref = "Normal")

# re-run DESeq:
design(dds_hp) <- formula(~ Haptoglobin_InflamStat + Parity_Classification + DIM_Collected + BCS + qpcr_16s_copies_ul)

dds_hp <- DESeq(dds_hp, 
                 test = "Wald", 
                 fitType = "parametric")
```

4.  DESeq Results and Log2 Fold Change Shrinkage

```{r DESeq_Results}
alpha <- 0.01
resultsNames(dds_hp)

res_hp <- results(dds_hp,
                   alpha = alpha,
                   name = "Haptoglobin_InflamStat_Elevated_vs_Normal")
summary(res_hp)

# Log2 Fold Change Shrink
res_LFC_hp <- lfcShrink(dds_hp, 
                         type = "ashr", 
                         res = res_hp)

# Filter for significant results
res_LFC_hp <- res_LFC_hp[which(res_LFC_hp$padj < 0.01 & 
                               abs(res_LFC_hp$log2FoldChange) > 2), ]
summary(res_LFC_hp)
        
# Order LFC results by adjusted P values
res_LFC_hp_df <- as.data.frame(res_LFC_hp[order(res_LFC_hp$padj, na.last = NA), ])
sum(res_LFC_hp_df$padj < 0.01, na.rm = TRUE)

# Keep only KOs with adjusted p-value < 0.01
res_LFC_hp_df <- res_LFC_hp_df[
  res_LFC_hp_df$padj < 0.01 & !is.na(res_LFC_hp_df$padj), ]
dim(res_LFC_hp_df)

# Save File
write.csv(res_LFC_hp, "DESeq2/Files/Results/Haptoglobin_InflamStat_full_model.csv")

# Plots
plotMA(res_hp, ylim = c(-30, 30))  # after DESeq2
plotMA(res_LFC_hp, ylim = c(-30, 30))  # after Log2 Fold Change shrinkage
```

5.  Volcano plot

```{r Hp_Volcano_plot}
library(ggplot2)
library(ggrepel)

# Add threshold labels
res_df <- res_LFC_hp_df
res_df$significant <- ifelse(res_df$padj < 0.01 & abs(res_df$log2FoldChange) > 1,
                             "Significant", "Not significant")

# 10 top genes
top_genes <- res_df[which(res_df$padj < 0.01 & abs(res_df$log2FoldChange) > 2), ]

# Volcano plot
ggplot(res_df, aes(x = log2FoldChange, y = -log10(padj))) +
  geom_point(aes(color = significant), alpha = 0.7, size = 2) +
  scale_color_manual(values = c("Significant" = "red", "Not significant" = "gray")) +
  geom_vline(xintercept = c(-1, 1), linetype = "dashed", color = "black") +
  geom_hline(yintercept = -log10(0.01), linetype = "dashed", color = "black") +
  labs(title = "Haptoglobin KOs - Volcano Plot",
       x = expression(Log[2]~Fold~Change),
       y = expression(-Log[10]~Adjusted~p~value)) + geom_text_repel(data = top_genes,
                  aes(label = rownames(top_genes)),
                  size = 3, box.padding = 0.3, max.overlaps = 10) + theme_minimal(base_size = 14)
```

```{r}
# Display table
head(res_LFC_hp_df)
```

## Anotation of KOS - Haptoglobin

```{r hp_annotation}
library(KEGGREST)

# KOs names:
KOs_names_Haptoglobin <- rownames(res_LFC_hp_df)
KOs_names_Haptoglobin <- gsub("ko:", "", KOs_names_Haptoglobin)
res_LFC_hp_df$X <- KOs_names_Haptoglobin


# Retrieve pathways for each KO
names(KOs_names_Haptoglobin) <- KOs_names_Haptoglobin

Haptoglobin_info <- lapply(KOs_names_Haptoglobin, function(ko) {
  ko_entry <- tryCatch(keggGet(paste0("ko:", ko))[[1]], error = function(e) NULL)
  ec_name <- if (!is.null(ko_entry) && !is.null(ko_entry$NAME)) ko_entry$NAME[1] else NA
  
  res <- tryCatch(keggLink("pathway", paste0("ko:", ko)), error = function(e) NULL)
  if (!is.null(res) && length(res) > 0) {
    path_ids <- sub("path:", "", res)
    path_names <- sapply(path_ids, function(pid)
      tryCatch(keggGet(pid)[[1]]$NAME, error = function(e) NA))
    list(EC_name = ec_name, Pathway = unique(na.omit(path_names)))
  } else {
    list(EC_name = ec_name, Pathway = NA)
  }
})

# Convert list to dataframe
Haptoglobin_info_df <- do.call(rbind, lapply(Haptoglobin_info, function(x) {
  # If multiple pathways, collapse into a single string separated by "; "
  pathway <- if (is.na(x$Pathway[1])) NA else paste(x$Pathway, collapse = "; ")
  data.frame(EC_name = x$EC_name, Pathway = pathway, stringsAsFactors = FALSE)
}))

#Update Data frame:
rownames(Haptoglobin_info_df) <- names(KOs_names_Haptoglobin)

Haptoglobin_merged <- cbind(res_LFC_hp_df,
                            Haptoglobin_info_df[res_LFC_hp_df$X, ])

write.csv(Haptoglobin_merged, "DESeq2/Files/Results/Haptoglobin_with_KEGG_info.csv")
```

```{r hp_annotation_table}
head(Haptoglobin_merged)
library(DT)

res_df_hp <- Haptoglobin_merged
res_df_hp$EC_name <- as.character(res_df_hp$EC_name)
res_df_hp$Pathway <- as.character(res_df_hp$Pathway)

colnames(res_df_hp)
head(rownames(res_df_hp))

if (!"KOs" %in% colnames(res_df_hp)) {
  res_df_hp <- cbind(KOs = rownames(res_df_hp), res_df_hp)
  rownames(res_df_hp) <- NULL
}


DT::datatable(
  res_df_hp,
  extensions = 'Buttons',
  options = list(
    dom = 'Bfrtip',
    buttons = c('copy', 'csv', 'excel', 'pdf'),
    pageLength = 10,
    lengthMenu = c(10, 15, 25, 50, 100),
    scrollX = TRUE
  ),
  filter = 'top',
  rownames = FALSE,
  class = 'stripe hover compact'
)

```

# Venn Diagram

```{r venn, message=FALSE, warning=FALSE}
library(VennDiagram)
library(dplyr)
padj_cutoff <- 0.01

# Fibrinogen
fibrinogen_up <- Fibrinogen_merged %>%
  filter(padj < padj_cutoff, log2FoldChange > 2) %>%
  pull(X) %>%
  unique()

fibrinogen_down <- Fibrinogen_merged %>%
  filter(padj < padj_cutoff, log2FoldChange < 2) %>%
  pull(X) %>%
  unique()

# Haptoglobin
haptoglobin_up <- Haptoglobin_merged %>%
  filter(padj < padj_cutoff, log2FoldChange > 2) %>%
  pull(X) %>%
  unique()

haptoglobin_down <- Haptoglobin_merged %>%
  filter(padj < padj_cutoff, log2FoldChange < 2) %>%
  pull(X) %>%
  unique()



# Plot
venn.plot <- venn.diagram(
  x = list(
    Fibrinogen_Up = fibrinogen_up,
    Fibrinogen_Down = fibrinogen_down,
    Haptoglobin_Up = haptoglobin_up,
    Haptoglobin_Down = haptoglobin_down
  ),
  filename = NULL,
  fill = c("#66c2a5", "#fc8d62", "#8da0cb", "#e78ac3"),
  alpha = 0.6,
  cex = 1.3,
  cat.cex = 1.2,
  cat.pos = 0,
  cat.dist = 0.05,
  margin = 0.1,
  main = "Overlap of Significant KOs by Condition and Direction"
)

grid::grid.newpage()
grid::grid.draw(venn.plot)
```
